Initial estuarine response to the nutrient-rich Piney Point discharge into Tampa Bay, Florida

### 2021-11-11 manu-draft.Rmd
Tweaks to Intro
Changing text to 'release', and updating some descriptors of potential gypstack impacts.

Aging infrastructure and climate change stressors acting on wastewater containment facilities are global concerns for the environmental protection of surface waters.  From March 30th to April 9th, 2021, 814 million liters of mixed legacy phosphate mining wastewater and marine dredge water from the Piney Point facility were released into Tampa Bay (Florida, USA).  An estimated 186 metric tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen load estimates in a matter of days.  In the period immediately after the discharge, elevated levels of phytoplankton (non-harmful diatoms) were observed in the nearshore regions of the lower Bay. Filamentous cyanobacteria blooms (*Dapis spp.*) peaked in June, followed by blooms of the red tide organism *Karenia brevis* that peaked in July.  Reported fish kills tracked *K. brevis* concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.  By comparing these observations to decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years, we conclude that all of these biological responses observed after the discharge from Piney Point are abnormal relative to historic conditions in Tampa Bay.

@@

Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources [@Tayibi09;@Hudson11]. Phosphate fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].  The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of phosphogypsum waste (CaSO$_4\cdot$ H$_2$O).  This waste is typically stored on-site in large earthen stacks (gypstacks) capable of holding millions of liters of process water.  Water quality in gypstacks can vary depending on processing method used at the mining facility, background geological characteristics of the region, and on-site practices for managing stormwater or other activities that can introduce additional materials to the holding ponds [@Henderson04;@Perez10].  In addition to elevated phosphorus concentrations, other nutrients, contaminants, and radionuclides may be present at values much higher than those of natural surface waters [@Burnett01;@Beck18b]. Many of these gypstacks no longer support active mining and aging infrastructure combined with climate change related stressors (e.g., heavy precipitation events) have reduced the capacity of these facilities to maintain water on site. Environmental and human health risks associated with these stacks can occur through controlled or uncontrolled discharge to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].

@@

The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer to support agricultural production in other countries [@Henderson04].  By 2001, an estimated 36 million metric tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].  Currently, seventeen phosphogypsum stacks exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.  The potential environmental impacts of these facilities are considered a critical threat that could halt or reverse the decades of prior ecosystem improvement. The Piney Point facility located in Palmetto, Florida is a large, remnant gypstack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [see supplement for a history of the facility, @Henderson04].   Holding capacity of the ponds has decreased over time from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee.  Discharges from the stacks occurred in the early 2000s and in 2011 to nearby Bishop Harbor connected to Tampa Bay.  Those discharges resulted in spatially-restricted, ecosystem responses including localized harmful algal blooms and increased macroalgal abundance [@Switzer11;@Garrett11].

@@

This paper provides an initial assessment of environmental conditions in Tampa Bay over six months following the recent release of legacy phosphate mining wastewater from March 30th to April 9th, 2021.  The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region. We focus on nitrogen inputs as the identified limiting nutrient for Tampa Bay and its potential to create water quality conditions unfavorable for seagrass growth due to enhanced algal production.  A timeline of events in 2021 is provided, which is supported by the quantitative results from 2021 response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency discharges from the Piney Point facility.  The results from this paper provide examples of anticipated short-term responses to acute nutrient loadings from legacy mining facilities in the broader context of historical conditions that may influence system response to these events.

@@

Monitoring of the natural resources of Tampa Bay in response to the discharge from Piney Point began in April, 2021 and continued for six months through September. These data were collected through a coordinated effort under the guidance of a discharge plume simulation by a numerical circulation model run by the Ocean Circulation Lab at the University of South Florida (USF), College of Marine Science.  The evolution of discharge from Piney Point was simulated using the Tampa Bay Coastal Ocean Model (TBCOM) nowcast/forecast system [@Chen18;@Chen19], with an embedded tracer module that included realistic discharge rates. Normalized tracer distributions were automatically updated each day, providing 1-day hindcasts and 3.5-day forecasts throughout the period of discharge and subsequent Tampa Bay distribution. The modeled plume evolution web product (Y. Liu, R.H. Weisberg, J. Chen, Y. Sun, personal communication) served as the principal guidance for coordinating the data collection during the event.

>>>>

Aging infrastructure and climate change stressors acting on wastewater containment facilities are global concerns for the environmental protection of surface waters.  From March 30th to April 9th, 2021, 814 million liters of mixed legacy phosphate mining wastewater and marine dredge water from the Piney Point facility were released into Tampa Bay (Florida, USA).  An estimated 186 metric tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen load estimates in a matter of days.  In the period immediately after the release, elevated levels of phytoplankton (non-harmful diatoms) were observed in the nearshore regions of the lower Bay. Filamentous cyanobacteria blooms (*Dapis spp.*) peaked in June, followed by blooms of the red tide organism *Karenia brevis* that peaked in July.  Reported fish kills tracked *K. brevis* concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.  By comparing these observations to decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years, we conclude that all of these biological responses observed after the release from Piney Point are abnormal relative to historic conditions in Tampa Bay.

@@

Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources [@Tayibi09;@Hudson11]. Phosphate fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].  The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of phosphogypsum waste (CaSO$_4\cdot$ H$_2$O).  This waste is typically stored on-site in large earthen stacks (gypstacks) capable of holding millions of liters of process water.  Water quality in gypstacks can vary depending on processing method used at the mining facility, background geological characteristics of the region, and on-site practices for managing stormwater or other activities that can introduce additional materials to the holding ponds [@Henderson04;@Perez10].  In addition to elevated phosphorus concentrations, other nutrients, contaminants, and radionuclides may be present at values much higher than those of natural surface waters [@Burnett01;@Beck18b]. Many of these gypstacks no longer support active mining and aging infrastructure combined with climate change related stressors (e.g., heavy precipitation events) have reduced the capacity of these facilities to maintain water on site. Environmental and human health risks associated with these stacks can occur through controlled or uncontrolled releases to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].

@@

The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer to support agricultural production in other countries [@Henderson04].  By 2001, an estimated 36 million metric tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].  Currently, seventeen phosphogypsum stacks exist in the Tampa Bay watershed.  Effective management and final closure of these facilities are imperative to reduce threats to prior ecosystem recovery efforts and investments. The Piney Point facility located in Palmetto, Florida is a large, remnant gypstack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [see supplement for a history of the facility, @Henderson04].   Holding capacity of the ponds has decreased over time from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee.  Unanticipated releases from the stacks occurred in the early 2000s and in 2011 to nearby Bishop Harbor connected to Tampa Bay.  Those releases resulted in spatially-restricted, ecosystem responses including localized harmful algal blooms and increased macroalgal abundance [@Switzer11;@Garrett11].

@@

This paper provides an initial assessment of environmental conditions in Tampa Bay over six months following the recent release of legacy phosphate mining wastewater from March 30th to April 9th, 2021.  The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region. We focus on nitrogen inputs as the identified limiting nutrient for Tampa Bay and its potential to create water quality conditions unfavorable for seagrass growth due to enhanced algal production.  A timeline of events in 2021 is provided, which is supported by the quantitative results from 2021 response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency releases from the Piney Point facility.  The results from this paper provide examples of anticipated short-term responses to acute nutrient loadings from legacy mining facilities in the broader context of historical conditions that may influence system response to these events.

@@

Monitoring of the natural resources of Tampa Bay in response to the release from Piney Point began in April, 2021 and continued for six months through September. These data were collected through a coordinated effort under the guidance of a plume simulation by a numerical circulation model run by the Ocean Circulation Lab at the University of South Florida (USF), College of Marine Science.  The plume evolution from Piney Point was simulated using the Tampa Bay Coastal Ocean Model (TBCOM) nowcast/forecast system [@Chen18;@Chen19], with an embedded tracer module that included realistic release rates. Normalized tracer distributions were automatically updated each day, providing 1-day hindcasts and 3.5-day forecasts throughout the period of discharge and subsequent Tampa Bay distribution. The modeled plume evolution web product (Y. Liu, R.H. Weisberg, J. Chen, Y. Sun, personal communication) served as the principal guidance for coordinating the data collection during the event.
